\documentclass[headsepline,footsepline,12pt,textcomp=full]{scrreprt}
\usepackage{scrhack}
\usepackage{csquotes}
\usepackage[hidelinks]{hyperref}
\usepackage{times}
\PassOptionsToPackage{full}{textcomp}


\usepackage{graphicx,color}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{verbatim}
\usepackage{longtable}
\usepackage{tabularx}       % Fuer Tabellen laenger als eine Seite
\usepackage{multicol}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{textcomp}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{newtxtext}
\usepackage{scrlayer-scrpage}
\usetikzlibrary{arrows.meta, positioning, shapes.geometric}
\usepackage[numbers,sort&compress]{natbib}

\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[english]{babel}
\usepackage{blindtext}
\usepackage[automark]{scrpage2}
% \pagestyle{scrheadings}
% Kopf- und Fußzeile
\clearscrheadfoot 
\chead{\headmark} % automatischen Kapitelnamen rein
\ofoot[\pagemark]{\pagemark} % oben rechts Seitenzahl laut Richtlinie
\ifoot{\invnr} % automatischen Kapitelnamen rein

\pagestyle{scrheadings}

% --------------------

% Seitendefinitionen

\setlength{\topmargin}{1.5cm}
\setlength{\headheight}{12pt}
\setlength{\headsep}{20pt}
\setlength{\topskip}{12pt}
\setlength{\evensidemargin}{0pt}
\setlength{\oddsidemargin}{0pt}
\setlength{\textheight}{240mm}
\setlength{\textwidth}{160mm}
\setlength{\voffset}{-2cm}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt}



\input{inc/profile.tex}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\invnr}{}

\begin{document}

\input{inc/title}
\pagenumbering{roman}
\tableofcontents
\listoffigures
\listoftables
\input{inc/abbreviations}
\newpage

\pagenumbering{arabic}
\input{inc/acknowlgement}
\input{inc/Abstract}

\input{inc/intro}
\input{inc/Literature Review}
\input{inc/methodology}
\input{inc/Result Section}
\input{inc/Conclusion and future work}

% Literaturverzeichniss
\appendix
\include{inc/appendix}
\newpage
% \bibliographystyle{alpha}
% % \bibliographystyle{apalike}
% \bibliography{bibs/litDB}
\bibliographystyle{IEEEtran}
\bibliography{bibs/litDB}


\newenvironment{declaration}
  {\cleardoublepage\chapter*{Declaration}%
   \addcontentsline{toc}{chapter}{Declaration}}
  {}
\begin{declaration}
\noindent I, \saauthor, hereby declare that the thesis entitled "VR Meditation:
Enhancing Relaxation through Adaptive
Audiovisual Cues and Biofeedback" and the work presented within it are entirely my own. 
I confirm that:

\begin{itemize}
    \item This work was carried out wholly or mainly during my candidature for a research degree at this University.
    
    \item Where any part of this thesis has previously been submitted for a degree or any other qualification at this or any other institution, this has been clearly stated.
    
    \item Where I have consulted the published work of others, this has been clearly attributed.
    
    \item Where I have quoted from the work of others, the source has always been given. Except for such quotations, this thesis is entirely my own work.
    
    \item I have acknowledged all major sources of assistance.
    
    \item Where the thesis includes work undertaken jointly with others, I have clearly indicated what was carried out by others and what I contributed myself.
\end{itemize}

\vspace{2em}

\noindent\textbf{Signed:}\\[1em]
\rule{25em}{0.5pt}

\vspace{2em}

\noindent\textbf{Date:}\\[1em]
\rule{25em}{0.5pt}

\end{declaration}


\end{document}



